---
title: "wilcoxon tests for pnps"
output: html_document
date: "2024-03-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, load and prep data, echo=FALSE, results='hide'}
rm(list = ls())
library(ggplot2)
library(ggpubr)

pnps_gw = read.delim("/users/sana/Documents/AIM3/021-full_pipeline/10-correction_pnps/all_pnps_corrected.tsv")
pnps_gw$Run = pnps_gw$SRA_id
pnps_gw$pNpS_genomewide = pnps_gw$pNpS

pnps_sgene = read.delim("/users/sana/Documents/AIM3/021-full_pipeline/10-correction_pnps/all_pnps_corrected_sgene.tsv")
pnps_sgene$Run = pnps_sgene$SRA_id


all_pnps = merge(pnps_gw[, c("Run", "pNpS_genomewide")],
                 pnps_sgene[, c("Run", "pNpS_sgene")],
                 "Run")

### filter based on sample quality:
all_data = read.delim("/users/sana/Documents/AIM3/021-full_pipeline/1-metrics/metrics_strict_thresholds_min50_5percent.tsv")
all_data = all_data[-which(all_data$mean_depth_from_bed < 50), ]
all_data = all_data[-which(all_data$min_50_site_count < 15000), ]


all_data$Species = relevel(as.factor(all_data$Species), ref = "human")

all_data$tissue_type = "swab"
all_data$tissue_type[which(all_data$tissue == "lymph-tissue")] = "lymph"
all_data$tissue_type = relevel(as.factor(all_data$tissue_type), ref = "swab")

all_data$lab_infected = relevel(as.factor(all_data$lab_infected), ref = "natural")
all_data$tissue = relevel(as.factor(all_data$tissue), ref = "nasal/pharyngeal-swab")
all_data$assay_type = relevel(as.factor(all_data$assay_type), ref = "AMPLICON")
all_data$BioProject = as.factor(all_data$BioProject)
all_data$shannon = as.numeric(all_data$filtered_shannon)
all_data$dpth = as.vector(scale(all_data$mean_depth_from_bed))
all_data$brdth = all_data$min_50_site_count/29903
all_data$filtered_richness = as.numeric(all_data$filtered_richness)
all_data$filtered_shannon = as.numeric(all_data$filtered_shannon)
all_data$Infection = all_data$lab_infected

### merge pnps with the rest of the data
all_data = merge(all_data, all_pnps, "Run")

swab_only = all_data[which(all_data$tissue_type == "swab"), ]

```


Shapiro test of normality for the distribution of pnps genomewide across *all samples*:
(pvalue < 0.05 means the distribution is significantly different from normal)
```{r, shapiro test 1, echo=FALSE}
shapiro.test(all_data$pNpS_genomewide)
```


Shapiro test of normality for the distribution of pnps genomewide on swab samples only:
```{r, shapiro test 2, echo=FALSE}
shapiro.test(swab_only$pNpS_genomewide)
```
**Histograms of the distribution of genomewide pnps, for all data and swab only samples:**
```{r, histograms of pnps genomewide, echo=FALSE}
ggplot(data = all_data) + geom_histogram(aes(x = pNpS_genomewide), color = "blue", fill = NA, binwidth = 0.01) + 
  theme_bw()+ggtitle("pnps genomewide - all data points")


ggplot(data = swab_only) + geom_histogram(aes(x = pNpS_genomewide), color = "blue", fill = NA, binwidth = 0.01) + 
  theme_bw() + ggtitle("pnps genomewide - swab only samples")

```


**Comparison of pnps in lymph vs swab samples, in deer samples:**

Genomewide pnps:
```{r, deer 1, echo=FALSE}
###### comparison of pnps genomewide and psgene in lymph and swab, in deer samples
deer_samples = all_data[which(all_data$Species == "deer"), ]

wilcox.test(deer_samples$pNpS_genomewide[which(deer_samples$tissue_type == "swab")],
            deer_samples$pNpS_genomewide[which(deer_samples$tissue_type == "lymph")])

print("mean in swab:")
mean(deer_samples$pNpS_genomewide[which(deer_samples$tissue_type == "swab")])
print("mean in lymph:")
mean(deer_samples$pNpS_genomewide[which(deer_samples$tissue_type == "lymph")])

```
(significant, mean lower in swab)


S gene: 
```{r, deer 2, echo=FALSE}
## for s gene
wilcox.test(deer_samples$pNpS_sgene[which(deer_samples$tissue_type == "swab")],
            deer_samples$pNpS_sgene[which(deer_samples$tissue_type == "lymph")])



```
(not significant)

**Comparison of pNpS across species for all samples**

1-1 : Genomewide on all data:
```{r, pnps comparison all samples, echo=FALSE}
results = data.frame(species1 = vector(),
                     species2 = vector(),
                     gene = vector(),
                     pvalue = vector(),
                     sample = vector())

species1 = "human"
for(species2 in c("cat", "dog", "mink", "deer")){
  t = wilcox.test(all_data$pNpS_genomewide[which(all_data$Species == species1)],
                  all_data$pNpS_genomewide[which(all_data$Species == species2)])
  
  results = rbind(results, 
                  data.frame(species1 = species1,
                             species2 = species2,
                             gene = "genomewide",
                             pvalue = t$p.value,
                             sample = "all data points")
                  )
  
  
  t = wilcox.test(all_data$pNpS_sgene[which(all_data$Species == species1)],
                  all_data$pNpS_sgene[which(all_data$Species == species2)])
  
   results = rbind(results, 
                  data.frame(species1 = species1,
                             species2 = species2,
                             gene = "S",
                             pvalue = t$p.value,
                             sample = "all data points")
                  )
  
}


print(results[which(results$gene == "genomewide"), ])

```

1-2: S gene on all data:
```{r, w2, echo=FALSE}
print(results[which(results$gene == "S"), ])
```


2-1: Genomewide on swab samples only:
```{r, wilcoxon on swab only samples, echo=FALSE}
results = data.frame(species1 = vector(),
                     species2 = vector(),
                     gene = vector(),
                     pvalue = vector(),
                     sample = vector())

species1 = "human"
for(species2 in c("cat", "dog", "mink", "deer")){
  t = wilcox.test(swab_only$pNpS_genomewide[which(swab_only$Species == species1)],
                  swab_only$pNpS_genomewide[which(swab_only$Species == species2)])
  
  results = rbind(results, 
                  data.frame(species1 = species1,
                             species2 = species2,
                             gene = "genomewide",
                             pvalue = t$p.value,
                             sample = "swab only points")
                  )
  
  
  t = wilcox.test(swab_only$pNpS_sgene[which(swab_only$Species == species1)],
                  swab_only$pNpS_sgene[which(swab_only$Species == species2)])
  
   results = rbind(results, 
                  data.frame(species1 = species1,
                             species2 = species2,
                             gene = "S",
                             pvalue = t$p.value,
                             sample = "swab only points")
                  )
  
}


print(results[which(results$gene == "genomewide"), ])


```

2-2: S gene on swab only samples:
```{r, w3, echo=FALSE}
print(results[which(results$gene == "S"), ])

```


when running the tests on swab only samples, deer changes slightly (the p value gets smaller) but not enough to be significant.



**The median and mean values for pnps in all samples:**
```{r, median values, echo=FALSE}

medians = data.frame(species = vector(),
                     gene = vector(),
                     median = vector(),
                     mean = vector(),
                     sample = vector())

for(species in c("cat", "dog", "mink", "deer", "human")){
  medians = rbind(medians,
                  data.frame(species = species,
                             gene = "genomewide",
                             median = median(all_data$pNpS_genomewide[which(all_data$Species == species)]),
                             mean = mean(all_data$pNpS_genomewide[which(all_data$Species == species)]),
                             sample = "all data points")
                  )
}

for(species in c("cat", "dog", "mink", "deer", "human")){
  medians = rbind(medians,
                  data.frame(species = species,
                             gene = "S",
                             median = median(all_data$pNpS_sgene[which(all_data$Species == species)]),
                             mean = mean(all_data$pNpS_sgene[which(all_data$Species == species)]),
                             sample = "all data points")
                  )
}

print(medians)

```

**The median and mean values for pnps in swab samples:**
```{r, means and medians in deer,echo=FALSE}
medians = data.frame(species = vector(),
                     gene = vector(),
                     median = vector(),
                     mean = vector(),
                     sample = vector())

for(species in c("cat", "dog", "mink", "deer", "human")){
  medians = rbind(medians,
                  data.frame(species = species,
                             gene = "genomewide",
                             median = median(swab_only$pNpS_genomewide[which(swab_only$Species == species)]),
                             mean = mean(swab_only$pNpS_genomewide[which(swab_only$Species == species)]),
                             sample = "swab only")
                  )
}

for(species in c("cat", "dog", "mink", "deer", "human")){
  medians = rbind(medians,
                  data.frame(species = species,
                             gene = "S",
                             median = median(swab_only$pNpS_sgene[which(swab_only$Species == species)]),
                             mean = mean(swab_only$pNpS_sgene[which(swab_only$Species == species)]),
                             sample = "swab only")
                  )
}

print(medians)


```


